**Sequencing 16S rRNA gene fragments using the PacBio SMRT DNA sequencing system**

### 2016-02-12 Schloss_PacBio16S_PeerJ_2016.Rmd
Add new PacBio reference

New advances in single molecule sequencing technologies are being developed to address this problem. One approach uses a random barcoding strategy to fragment, sequence, and assemble full-length amplicons using Illumina's HiSeq platform [@Burke2014; @Miller2013]. Although the algorithms appear to have minimized the risk of assembly chimeras, it is unclear what the sequencing error rate is by this approach. An alternative is the use of single molecule technologies that offer read lengths that are thousands of bases long. Although the Oxford Nanopore Technology has been used to sequence 16S rRNA genes [@BentezPez2016], the the platform produced by Pacific Biosciences (PacBio) has received wider attention for this application [@Mosher2013; @Mosher2014; @Fichot2013; @Schloss2015]. The PacBio Single Molecule, Real-Time (SMRT) DNA Sequencing System ligates hairpin adapters (i.e. SMRTbells) to the ends of double-stranded DNA. Although the DNA molecule is linear, the adapters effectively circularize the DNA allowing the sequencing polymerase to process around the molecule multiple times [@Au2012]. According to Pacific Biosciences the platform is able to generate median read lengths longer than 8 kb with the P6-C4 chemistry; however, the single pass error rate is approximately 15%. Given the circular nature of the DNA fragment, the full read length can be used to cover the DNA fragment multiple times resulting in a reduced error rate. Therefore, one should be able to obtain multiple coverage of the full 16S rRNA gene at a reduced error rate.

Despite the opportunity to potentially generate high-quality full-length sequences, the Pacific Biosciences platform has not been widely adopted for sequencing 16S rRNA genes [@Fichot2013; @Mosher2014; @Mosher2013]. Previous studies utilizing the technology have removed reads with mismatched primers and barcodes, ambiguous base  calls, and low quality scores [@Fichot2013]. Others have utilized the platform without describing the bioinformatic pipeline that was utilized [@Mosher2014; @Mosher2013]. The only study to report the error rate of the platform for sequencing 16S rRNA genes with a mock community used the P4-C2 chemistry and obtained an error rate of 0.32%, which is 16-fold higher than has been observed using the MiSeq or 454 platforms [@Schloss2011a; @Kozich2013]. In the current study, we assessed the quality of data generated by the PacBio sequencer using the improved P6-C4 chemistry and on-sequencer data processing. The goal was to determine whether this strategy could fill the need for generating high-quality, full-length sequence data on par with other platforms. We hypothesized that by modulating the 16S rRNA gene fragment length we could alter the read depth and obtain reads longer than are currently available by the 454 and Illumina platforms but with the same quality. To test this hypothesis, we developed a sequence curation pipeline that was optimized by reducing the sequencing error rate of a mock bacterial community with known composition. The resulting pipeline was then applied to 16S rRNA gene fragments that were isolated from soil and human and mouse feces.

>>>>

New advances in single molecule sequencing technologies are being developed to address this problem. One approach uses a random barcoding strategy to fragment, sequence, and assemble full-length amplicons using Illumina's HiSeq platform [@Burke2014; @Miller2013]. Although the algorithms appear to have minimized the risk of assembly chimeras, it is unclear what the sequencing error rate is by this approach. An alternative is the use of single molecule technologies that offer read lengths that are thousands of bases long. Although the Oxford Nanopore Technology has been used to sequence 16S rRNA genes [@BentezPez2016], the the platform produced by Pacific Biosciences (PacBio) has received wider attention for this application [@Mosher2013; @Mosher2014; @Fichot2013; @Schloss2015; @Singer2016]. The PacBio Single Molecule, Real-Time (SMRT) DNA Sequencing System ligates hairpin adapters (i.e. SMRTbells) to the ends of double-stranded DNA. Although the DNA molecule is linear, the adapters effectively circularize the DNA allowing the sequencing polymerase to process around the molecule multiple times [@Au2012]. According to Pacific Biosciences the platform is able to generate median read lengths longer than 8 kb with the P6-C4 chemistry; however, the single pass error rate is approximately 15%. Given the circular nature of the DNA fragment, the full read length can be used to cover the DNA fragment multiple times resulting in a reduced error rate. Therefore, one should be able to obtain multiple coverage of the full 16S rRNA gene at a reduced error rate.

Despite the opportunity to potentially generate high-quality full-length sequences, it is surprising that the Pacific Biosciences platform has not been more widely adopted for sequencing 16S rRNA genes. Previous studies utilizing the technology have removed reads with mismatched primers and barcodes, ambiguous base  calls, and low quality scores [@Fichot2013] or screened sequences based on the predicted error rate [@Singer2016]. Others have utilized the platform without describing the bioinformatic pipeline that was utilized [@Mosher2014; @Mosher2013]. The only study to report the error rate of the platform for sequencing 16S rRNA genes with a mock community used the P4-C2 chemistry and obtained an error rate of 0.32%, which is 16-fold higher than has been observed using the MiSeq or 454 platforms [@Schloss2011a; @Kozich2013]. In the current study, we assessed the quality of data generated by the PacBio sequencer using the improved P6-C4 chemistry and on-sequencer data processing. The goal was to determine whether this strategy could fill the need for generating high-quality, full-length sequence data on par with other platforms. We hypothesized that by modulating the 16S rRNA gene fragment length we could alter the read depth and obtain reads longer than are currently available by the 454 and Illumina platforms but with the same quality. To test this hypothesis, we developed a sequence curation pipeline that was optimized by reducing the sequencing error rate of a mock bacterial community with known composition. The resulting pipeline was then applied to 16S rRNA gene fragments that were isolated from soil and human and mouse feces.
